---
title: "Assignment3"
output: html_document
date: "2024-02-09"
author: "Semyon Toybis"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

We are examining the list of college majors from the FiveThiryEight article [The Economic Guide To Picking A College Major](https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/).

Below we import the data from the FiveThirtyEight Github

```{r cars}

library(tidyverse)
library(readr)

fileURL = 'https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv'

majorsDF = read.csv((url(fileURL)))
```

We are interested in finding majors that contain the phrases "DATA" or "STATISTICS"

```{r}

majorsOfInterest <- grep("DATA|STATISTICS", majorsDF$Major, value = TRUE, ignore.case = FALSE)

print(majorsOfInterest)
```

# Question 2

We are interested in transforming the below vector into one line of output that is separated by commas

```{r}

foodVector <- c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")

str_flatten_comma(foodVector)
```

# Question 3

Describe, in words, what these expressions will match:

-   (.)\\1\\1

    -   find a string that has the same charter three times in a row

    -   none of the strings in fruit or words fill the requirement

        ```{r}
        str_view(fruit, "(.)\\1\\1")
        str_view(words, "(.)\\1\\1")

        str_view(c('abbba', 'goooal', 'ball'), "(.)\\1\\1") # returns only 'abbba' and 'goooal' with the repeating characters highlighted
        ```

-   "(.)(.)\\\\2\\\\1"

    -   find a string that has two characters followed by the same two characters but in reverse order

        ```{r}
        str_view(words, "(.)(.)\\2\\1")
        ```

-   (..)\\1

    -   find a string that has the same two characters repeating

        ```{r}
        str_view(words,"(..)\\1")
        str_view(fruit,"(..)\\1")
        ```

-   "(.).\\\\1.\\\\1"

    -   find a string that has a character, followed by a different character, followed by the original character, followed by a different character, followed by the original character

        ```{r}
        str_view(words,"(.).\\1.\\1")
        str_view(fruit,"(.).\\1.\\1")
        ```

-   "(.)(.)(.).\*\\\\3\\\\2\\\\1"

    -   finds a string where there are three characters followed by any set of characters followed by the aforementioned three characters in reverse order

        ```{r}
        str_view(words,"(.)(.)(.).*\\3\\2\\1")
        ```

# Question 4

Construct regular expressions to match words that:

-   Start and end with the same character.

    ```{r}
    str_view(words, "^(.).*\\1$")
    ```

-   Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)

    ```{r}
    str_view(words, "(..).*\\1")
    ```

-   Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)

    ```{r}
    str_view(words, "(.).*\\1.*\\1")
    ```
