---
title: "assignment4"
author: "Semyon Toybis"
date: "2024-02-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Importing data

I created a csv file that has the data provided for the assignment and uploaded it to my github. I pull the data from github into a data frame, to allow for reproduction.

```{r}
library(tidyverse)
library(readr)

fileURL = 'https://raw.githubusercontent.com/stoybis/DATA607Repo/main/assignment4data.csv'

flightData = read.csv(url(fileURL), header = TRUE)

head(flightData)
```

## Cleaning the data

I drop the blank third row and assign column names to the first two columns

```{r}

#drop blank third row
flightData <-flightData[-3,]
rownames(flightData) <-NULL #reset the row names

#add column names to first two columns
colnames(flightData)[1:2] <- c('airline', 'status')

head(flightData)
```

I will be pivoting the data frame to a longer format. However, there are missing values in the data frame in row 2 column 1 and row 4 column 1. I fill in the blanks before pivoting.

```{r}
#fill in missing data
flightData[2,1] <- "ALASKA"
flightData[4,1] <- "AM WEST"

head(flightData)
```

## Pivoting the data frame

I want to pivot the data frame into a longer format where the columns are: airline, status, city, and count of flights.

```{r}
flightDataLong <- flightData |> pivot_longer(
  cols = !c('airline', 'status'),
  names_to = 'city',
  values_to = 'count'
)

head(flightDataLong)
```

I want to create two columns, on_time and delayed. This will allow me to sum across the two columns to get the total flights for each city by each airline. I rename "on time" to "on_time" to remove the space in the variable name

```{r}
flightDataLong <- flightDataLong |> pivot_wider(
  names_from = status,
  values_from = count
)

colnames(flightDataLong)[3] <- 'on_time'

head(flightDataLong)
```

## Mutations

I add two columns: total flights and delay rate which is equal to delays divided by total flights.

```{r}
flightDataLong <- flightDataLong |> mutate(total = on_time + delayed,
                                             delay_rate = delayed/total)

head(flightDataLong)
```

## Analysis

Below, I check to see which airline has the higher delay rate by each city.

```{r}
ggplot(flightDataLong, aes(x = city, y = delay_rate, fill = airline))+
  geom_bar(stat='identity', position = 'dodge') +
  ggtitle('Delay Rate by City')
```

The chart shows that AM West has a higher delay rate than Alaska for each city. The highest delay rates are in San Francisco while the lowest delay rates are in Phoenix.

## Simpson's Paradox

If we look at the total on time and delayed departures without breaking it down by city, AM West will seem to have less delays.

```{r}
totalFlights <- flightDataLong |> group_by(airline) |> summarize(total_onTime = sum(on_time), total_Delayed = sum(delayed))

totalFlights <- totalFlights |> mutate(totalFlights = total_onTime + total_Delayed)

totalFlights <- totalFlights |> mutate(delay_rate = total_Delayed / totalFlights)


ggplot(totalFlights, aes(x = airline, y = delay_rate, fill = airline))+
  geom_bar(stat='identity', position = 'dodge') +
  ggtitle('Delay Rate by Airline')
```
